### 2016-03-01 paper/icfp16-submission/semantics.tex
swap order of lemmas 2 and 3

\paragraph{The Type of a a Value} The (dynamic) type of a
value $v$ is defined as a function $\typeof{v}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{\thole}} & \defeq & \thole \\
    % \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of values rather than
performing full type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
instead of the usual $\tint \rightarrow \tint$.

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$'s input, \ie a partial instantiation of \thole.
%
We can extract this type from an evaluation trace.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ upto $\trace$, written
\ptype{\trace}{f}, is
$\typeof{\subst{\vsu}{\ehole}}$.
\ES{should we say $\subst{\tsu}{\thole}$ instead? would need a helper function that does the repeated application of \tsu\ until the result has no $\thole \in dom(\tsu)$}
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial input type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$ (\ie $\trace'$ is a single-step
extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we may instantiate the hole with
  a concrete value.
  %
  But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a successful call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ invokes \forcesym.
\end{lem}

\begin{proof}
  By case analysis on the evaluation rules.
  %
  If $\ptype{\trace}{f} \neq \ptype{\trace}{f}$ then one of the holes in $f$'s
  argument must have been instantiated with a concrete value at the last step.
  %
  An examination of the rules shows that only place this happens is
  in the second case of \forcesym.
\end{proof}

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $s_{\trace} = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},...,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}
  in at most $k$ steps:
%
  \begin{description}
  \item [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  \end{description}
\end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},...,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \subst{\vsu}{\ehole}$ demonstrates that
  $\eapp{f}{v}$ gets stuck. \ES{here we also need a repeated application of the substitution, as \ehole may map directly to $\vnode{\thole}{\vhole{\thole}_1}{\vleaf{\thole}}{\vleaf{\thole}}$ and $\ehole_1$ may later map to \etrue.}
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
  such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
\end{description}
\end{proof}

>>>>

\paragraph{The Type of a a Value} The (dynamic) type of a
value $v$ is defined as a function $\typeof{v}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{\thole}} & \defeq & \thole \\
    % \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of values rather than
performing traditional type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
instead of the usual $\tint \rightarrow \tint$.

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$'s input, \ie a partial instantiation of \thole.
%
We can extract this type from an evaluation trace, by repeatedly
applying the final type substitution to \thole until it contains no
holes in the domain of the substitution. We will call this process of
repeated substition \emph{resolving} a hole, and will use
\resolve{\thole}{\tsu} to denote the the resolution of \thole with
respect to \tsu.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ upto $\trace$, written
\ptype{\trace}{f}, is
$\resolve{\thole}{\tsu}$.
%$\typeof{\subst{\vsu}{\ehole}}$.
% \ES{should we say $\subst{\tsu}{\thole}$ instead? would need a helper function that does the repeated application of \tsu\ until the result has no $\thole \in dom(\tsu)$}
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

% \begin{lem}
% \label{lem:narrow-tsu}
% If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
% and $\trace' \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
%     (\ie $\trace'$ is a single-step extension of $\trace$)
% and $\tsu \neq \tsu'$
% then the final step must have \emph{successfully} invoked \forcesym.
% \end{lem}

\paragraph{Narrowing}
%
Only a successful call to \forcesym can change the partial
input type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$
(\ie $\trace'$ is a single-step extension of $\trace$)
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ invokes \forcesym.
\end{lem}

\begin{proof}
  By case analysis on the evaluation rules.
  %
  If $\ptype{\trace}{f} \neq \ptype{\trace'}{f}$ then,
  % one of the holes in $f$'s
  % argument must have been instantiated with a concrete value at the last step.
  by the definition of $\ptype{\trace}{f}$, $\tsu \neq \tsu'$, as \thole
  does not change.
  %
  % An examination of the rules shows that only place this happens is
  % in the second case of \forcesym.
  An examination of the rules shows that only \forcesym can update \tsu,
  and furthermore that only the successful cases of \forcesym do update
  \tsu.
\end{proof}

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial input type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
$\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  First note that by Lemma~\ref{lem:force-inst} we can immediately
  discharge the \rulename{E-*-Bad} rules as they cannot change
  \ptype{\trace}{f} at all, and are thus trivially
  compatibility-preserving. For the \rulename{E-*-Good} rules we can
  show that, by virtue of \forcesym succeeding, all must preseve
  compatibility.
  % Note that all rules preserve partial types with the exception of when
  % \forcesym\ is called on a hole, in which case we may instantiate the hole with
  % a concrete value.
  % %
  % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
\end{proof}

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial input type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
     in at most $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$. Let $s_{\trace} = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
  in at most $k$ steps.
%
  \begin{description}
  \item [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    Since $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$ we know
    that $s_{\trace'} \neq s_{\trace}$, and
    by Lemma~\ref{lem:force-inst} we know that we must have
    successfully called \forcesym at step $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the evaluation rules shows that narrowing to
    $t$ instead cannot succeed.
  \end{description}
\end{proof}

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t = \ptype{\trace}{f}$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with $t$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \resolve{\ehole}{\vsu}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
  such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
  % \ES{do we need to say anythign else?}
\end{description}
\end{proof}
